
\begin{thebibliography}{99}
  
\bibitem{gurgen} Gürgen, S. F., S. Yümlü., N. Okay,
  {\em A comparison of global, recurrent and smoothed-piecewise neural models
    for Istanbul stock exchange (ISE) prediction}, 
  Pattern Recognition Letters, Volume 26, Issue 13, Pages 2093-2103, 
  1 October 2005.

\bibitem{rabiner} Rabiner L. R. , 
  {\em A Tutorial on Hidden Markov Models and Selected Applications in Speech Recognition}, 
  Proceedings of IEEE vol. 77, no. 2, pp. 257-286, 
  1989.

\bibitem{ghahramani} Roweis S. and Z. Ghahramani,
  {\em  A Unifying Review of Linear Gaussian Models },
   Neural Computation 11(2):305--345, 
   1999.
   
\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{alpaydin} Alpaydın E., 
  {\em Introduction to Machine Learning},
  MIT Press, 
  2004.   
   
\bibitem{ghahramani3} Ghahramani Z., H. E. Hinton,
  {\em Switching State Space Models},
  Technical Report CRG-TR-96-3, Dept. Comp. Sci., Univ. Toronto, 
  1996.
   
\bibitem{shumway} Shumway R., H. S. Stoffer
  {\em Time series analysis and its applications 2nd Edition},
   New York, Springer, (Springer texts in statistics),
   2000.
   
\bibitem{jordan} Jordan M. I. , C. Bishop
  {\em An Introduction to Graphical Models},
   Not yet published, 
   2000.

\bibitem{kalman} Kalman R. E.,
  {\em A New Approach to Linear Filtering and Prediction Problems},
  Transactions of the ASME-Journal of Basic Engineering, 82 (Series D): 35-45, 
  1960.

\bibitem{kalman2} Kalman, R.E. and R.S. Bucy, 
  {\em New results in filtering and prediction theory},
  Trans. ASME J. Basic Eng., 83, 95-108,
  1961.
  
\bibitem{welling} Welling, M.,
  {\em The Kalman Filter - Lecture Tutorial},
   California Institute of Technology,
   2008.
   
\bibitem{lall} Lall, S.,
  {\em Modern Control 2 Lecture Notes},
   Stanford University,
   2006.
   
\bibitem{fama} Fama, E. F.,
  {\em The behavior of stock-market prices},
  The Journal of Business of the University of Chicago, 38, no. 1.
  1965.

\bibitem{famar} Fama, E. F.,
  {\em Random walks in stock market prices,},
  Financial Analysts Journal, 21 (5): 55-59.,
  1965.
  
\bibitem{tsay} Tsay, R. S.,
  {\em Time Series and Forecasting: Brief History and Future Research,},
  Journal of the American Statistical Association, Vol. 95, No. 450., pp. 638-643.,
  Jun 2000.

\bibitem{bachelier} Bachelier L.,
  {\em Thorie de la speculation. Annales Scientiques de l'Ecole Normale Superieure, III-17:21-86}
  Thesis for the Doctorate in Mathematical Science., 1900
  Reprinted by Editions Jacques Gabay, Paris 1995. English
  translation in The Random Character of Stock Market Prices, Ed. P. Cootner, pp
  1778, Cambridge, MIT Press, 1964.

\bibitem{shi} Shi S. and A. S. Weigend., {\em Taking Time Seriously: Hidden
  Markov Experts Applied to Financial Engineering}. In Proceedings of the
  IEEE/IAFE 1997 Conference on Computational Intelligence for Financial
  Engineering , pages 244-252. IEEE, 
  1997.

\bibitem{hassan} Hassan Md. R., Baikunth Nath, 
  {\em Stock Market Forecasting Using Hidden Markov Model: A New Approach},
  192-196, 
  ISDA 2005.

\bibitem{nabney} Azzouzi M.  and Ian T. Nabney,
  {\em Modelling Financial Time Series with Switching State Space Models},
  \PVerb!citeseer.ist.psu.edu/267552.html!,
  1999.
  
\bibitem{engle} Engle, Robert F. 
  {\em GARCH 101: The Use of ARCH/GARCH Models in Applied Econometrics},
  Journal of Economic Perspectives-Volume 15, Number 4-Pages 157-168, Fall
  2001. 
  
\bibitem{engle2} Engle, R. F. , Victor K. Ng,  
  {\em Measuring and Testing the Impact of News on Volatility},
  The Journal of Finance, Volume XLVIII, No. 5, 
  December 1993.
  
\bibitem{engle3} Engle, Robert F., {\em Autoregressive Conditional
  Heteroskedasticity with Estimates of the Variance of United Kingdom
  Inflation}, Econometrica. 50:4, pp. 987-1007. University of California, San
  Diego. Revision,
  2002.

\bibitem{bollershev} Bollershev, T.,
  {\em Generalized autoregressive conditional heteroskedasdicity}, Journal of
  Econometrics 31, 307-327,
  1986.
  
\bibitem{nelson} Nelson D., 
  {\em Conditional heteroskedasticity in asset returns: A new approach},
  Econometrica, 59, 347-370, 
  1990.
    
\bibitem{mandelbrot} Mandelbrot B. B., 
  {\em The variation of certain speculative prices, Journal of Business},
  XXXVI (1963), pp. 392-417, 
  1963.

  
\end{thebibliography}

